---
title: "Growth Profile Formatting"
author: "Elena"
date: "1/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(here)
library(tidyverse)
library(ggpubr)
library(viridis)
```

## Growth Profile Data

This document outlines the steps used to format and combine the growth profile data. Code from this document was used to create  growthProfiles.R, which is called in all other .Rmd files to read in the growth data. 

Guide:
  allGE = raw growth data from experiments GE2,3,4 and 5 (GE = "growth experiment")
  cleanGExMaxOD = all growth data with problematic replicates removed, maxOD calculated
  dODfinal = all reps with change in OD from t=0 to time at max OD
  dODwide = wide version of dODfinal
  tax = strain taxonomy info

```{r}
# import and organize data
ge2 <- read_excel(here('data', 'ge2_all copy.xlsx')) %>% mutate(exp = 'ge2')
ge3 <- read_excel(here('data', 'ge3_all copy.xlsx')) %>% mutate(exp = 'ge3')
ge4 <- read_excel(here('data', 'ge4_all copy.xlsx')) %>% mutate(exp = 'ge4')
ge5 <- read_excel(here('data', 'ge5_all copy.xlsx')) %>% mutate(exp = 'ge5')
tax <- read_excel(here('data', 'strain_tax copy.xlsx'))
# note that I added an entry in the strain_tax excel file so that the negative controls would NOT be removed when inner joining

# combine data, make variables consistent
allGE <- bind_rows(ge2, ge3, ge4, ge5) %>% 
  dplyr::select(sample, plate_id, well, tmpt, OD600, medium, strain, exp) %>%
  mutate(strain = replace(strain, strain %in% c('none', 'blank'), 'negative')) %>% #rename neg ctrls, remake grp to include negative
  unite('grp', medium, strain, sep = '.', remove = FALSE) %>%
  unite('grp2', exp, grp, sep = '.', remove = FALSE)
rm(ge2, ge3, ge4, ge5)

# add taxonomy info to allGE
allGE <- allGE %>% inner_join(tax, by = 'strain') 
# make the media factors to order data in plots
allGE$medium <- factor(allGE$medium, 
                       levels = c('difcoMB', 'HMBcmpt', 'HMBpep', 'HMBaa', 'HMBlips', 'HMBoligo', 
                                  'HMBorg', 'HMBntrl', 'HMBamisug', 'HMBacdsug', 'HMB--'))

# clean up the data in allGE
cleanGE <- allGE %>%
  # remove PR1white and PR1red, did not finish resequencing to identify
  filter(!strain %in% c('PR1red', 'PR1white')) %>%
  # I made a mistake loading these samples, see lab notebook
  filter(!(strain %in% c('atlcs', 'fors') & exp == 'ge3')) %>%
  # remove aberrant time points, OD spikes/valleys likely to be technical artifacts 
  filter(!(grp %in% c('HMBcmpt.bdim', 'HMB--.bdim') & tmpt == 162)) %>%
  filter(!(grp == 'HMBacdsug.fang' & tmpt == 189)) %>%
  filter(!(grp %in% c('HMBntrl.kuro', 'HMBamisug.kuro', 'HMBacdsug.kuro') & tmpt == 189))

# list of strains to loop through
clean_str <- unique(cleanGE$strain) %>% sort()
# don't include no-cell control in loop
clean_str <- clean_str[!clean_str == 'negative']
```

```{r}
# find maxOD, the timepoint at which the sample achieved the highest average OD600
# find the average maxOD across technical replicates, and the time point at which it's reached
maxOD <- cleanGE %>% 
  filter(tmpt < 265) %>% # only 2 experiments went longer than 264 hours
  group_by(grp2, tmpt) %>% mutate(avgMaxOD = mean(OD600)) %>% 
  dplyr::select(grp2, tmpt, avgMaxOD) %>% 
  ungroup() %>% group_by(grp2) %>% filter(avgMaxOD == max(avgMaxOD)) %>% distinct() %>% rename(t_maxOD = tmpt)

# some of the strains maintain the max OD for several time points, for these strains, I'm going to take the first time point that the average max OD was reached
# the maxOD df has the first time point at which the average max OD was reached for every strain
maxOD <- maxOD %>% group_by(grp2) %>% filter(t_maxOD == min(t_maxOD)) 

# add the t_maxOD and avgMaxOD to each row of cleanGE
# this df is useful for plotting the maxOD line
cleanGEwMaxOD <- cleanGE %>% inner_join(maxOD, by = 'grp2')

# to get all the dODs for each sample, I need to find the OD at t=0 and OD at t=maxOD
# create df with just the t_maxOD
# now I have OD for each well at time of average maxOD
dOD <- cleanGEwMaxOD %>% filter(tmpt == t_maxOD) %>% 
  dplyr::select(plate_id, well, grp2, exp, grp, medium, strain, t_maxOD, OD600, avgMaxOD) %>%
  rename(ODatMaxAvg = OD600)
# create a second df with just the info at t=0
ODt0 <- cleanGE %>% filter(tmpt == 0) %>% dplyr::select(grp2, 'plate_id', well, OD600) %>% rename(ODt0 = OD600)
# combine the info for maxOD and t=0, calculate the change in OD for each between t=0 and the OD at the time at which average OD was highest
dOD <- dOD %>% inner_join(ODt0, by = c('grp2', 'plate_id', 'well')) %>% 
  mutate(dOD = ODatMaxAvg - ODt0) %>% ungroup() %>%
  group_by(grp) %>% mutate(dODavgExp = mean(dOD), sdODavgExp = sd(dOD)) 

rm(ODt0, maxOD)
```

## Determine positive growth for binary and "adjusted" growth profiles

```{r}
# KW test for differences btw strains on each medium
dOD %>%
  ggplot(aes(x = strain, y = dOD)) +
  geom_boxplot() +
  facet_wrap(~medium, ncol = 4) +
  stat_compare_means(method = "kruskal.test", label.y = 1.8, label.x = 10) +
  guides(color = "none") +
  labs(title = 'Significant differences between dOD of strains grown on same medium') +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1), text = element_text(size = 6),
        legend.position = 'bottom')
```

```{r}
# Dunn's test for all pairs of strains on each medium
media <- unique(dOD$medium)
media.dunn <- tibble()

for (m in media){
  dat <- dOD %>% filter(medium == m)
  med.dn <- FSA::dunnTest(dOD ~ strain, dat, method = 'none', two.sided = TRUE)
  x <- med.dn[[2]] %>% separate(Comparison, c("strain1", "strain2"), sep = ' - ') %>% mutate(medium = m)
  media.dunn <- bind_rows(media.dunn, x)
}

# select only the comparisons with the negative control
s1neg <- media.dunn %>% 
  filter(strain1 == 'negative') %>% 
  dplyr::select(medium, strain2, Z, P.unadj) %>%
  rename(strain = strain2)
s2neg <- media.dunn %>% 
  filter(strain2 == 'negative') %>% 
  dplyr::select(medium, strain1, Z, P.unadj) %>%
  rename(strain = strain1)
media.sigVsNeg <- bind_rows(s1neg, s2neg)

rm(m, dat, med.dn, x, s1neg, s2neg, media.dunn)
```

```{r}
# correct p.values for multiple comparisons
media.adjPval <- p.adjust(media.sigVsNeg$P.unadj, method = "BH")
media.sigVsNeg$P.adj <- media.adjPval

dOD.pval <- dOD %>% 
  ungroup() %>% 
  dplyr::select(grp, medium, strain) %>% 
  distinct() %>% 
  inner_join(media.sigVsNeg, by = c('medium', 'strain')) %>%
  mutate(growth = ifelse(P.adj < .05, 'yes', 'no')) %>%
  dplyr::select(grp, growth)

# write p-values to file
x <- dOD %>% 
  ungroup() %>% 
  dplyr::select(medium, strain, dODavgExp) %>% 
  distinct() %>% 
  inner_join(media.sigVsNeg, by = c('medium', 'strain')) 
#writexl::write_xlsx(x, "pvalsGrowth.xlsx")

rm(x)
rm(media.adjPval)
```

```{r}
# create a new df in which strains that don't grow have dOD set to zero
dODfinal <- dOD %>% 
  inner_join(dOD.pval, by = 'grp') %>% 
  dplyr::select(exp, grp2, grp, strain, medium, dOD, growth) %>%
  mutate(dODadj = ifelse(growth == 'no', 0, dOD)) %>%
  mutate(dODadj = ifelse(is.na(dODadj), 0, dODadj))

# psych and jeju are the only two strains that do not grow on any medium 
dODfinal %>%
  ungroup() %>%
  filter(!strain == "negative" & growth == "no") %>%
  select(strain, medium) %>%
  distinct() %>%
  group_by(strain) %>%
  mutate(num = n()) %>%
  filter(num == 11) %>%
  select(strain) %>% distinct()

# remove jeju and psych, get mean of adjusted dODs
dODfinal <- dODfinal %>% 
  filter(!strain %in% c("jeju", "psych")) %>%
  group_by(strain, medium) %>% 
  mutate(deltaOD = mean(dODadj)) %>% 
  ungroup()

# spread data 
deltaOD <- dODfinal %>% 
  dplyr::select(strain, medium, deltaOD) %>% distinct() %>%
  pivot_wider(names_from = medium, values_from = deltaOD) %>%
  dplyr::select(strain, difcoMB, HMBcmpt, HMBpep, HMBaa, HMBlips, HMBoligo, 
                HMBorg, HMBntrl, HMBamisug, HMBacdsug, `HMB--`) %>%
  ungroup()



rm(dOD.pval, media.sigVsNeg, cleanGEwMaxOD, media, clean_str, cleanGE)
```

## Create growth curves for figure 1

```{r}
#install.packages("rcartocolor")
library(rcartocolor)
# The goal of rcartocolor is to provide color schemes for maps and other graphics designed by CARTO as described at https://carto.com/carto-colors/.
```

```{r}
### figure 1B and 1C ###
# use alteromonas 26499 as the example strain for growth curves in figure 1
str = "26499"

a <- allGE %>%
  filter(strain == str) %>%
  filter(tmpt < 265) %>%
  filter(exp == 'ge5') %>%
  select(strain, medium, plate_id, tmpt, OD600) %>%
  group_by(plate_id, medium, tmpt) %>%
  mutate(avgOD = mean(OD600), sdOD = sd(OD600)) %>%
  ggplot(aes(x = tmpt, y = avgOD, color = medium)) +
  geom_point() +
  geom_errorbar(aes(ymin = avgOD-sdOD, ymax = avgOD+sdOD), width = 7.5) +
  geom_line() +
  scale_color_manual(values = c("#A5AA99", "#5D69B1", "#52BCA3", "#99C945", "#CC61B0",
                                "#24796C","#DAA51B", "#2F8AC4", "#764E9F", "#ED645A",
                                "#E58606")) + 
  facet_wrap(~medium, nrow = 1) +
  theme_classic() +
  theme(legend.position = "none",
        text = element_text(size = 14)) +
  labs(x = 'Time (hours)',
       y = 'OD600 MED64')

b <- dODfinal %>%
  dplyr::select(strain, medium, deltaOD) %>% distinct() %>%
  ggplot(aes(x = medium, y = deltaOD, color = medium)) +
  geom_boxplot(outlier.size = -1) +
  geom_jitter(size = 2, alpha = .5, width = .2) +
  theme_classic() +
  theme(legend.position = "none",
        text = element_text(size = 14)) +
  labs(y = 'Max Change in OD600',
       x = "Medium") +
  scale_color_manual(values = c("#A5AA99", "#5D69B1", "#52BCA3", "#99C945", "#CC61B0",
                                "#24796C", "#DAA51B", "#2F8AC4", "#764E9F", "#ED645A",
                                "#E58606")) 


# curves and boxplot 
cowplot::plot_grid(a, b, ncol = 1, rel_heights = c(1, 2), labels = c("B", "C"))
```

### Number of strains per medium

```{r}
deltaODbin <- select(deltaOD, -strain)
deltaODbin[deltaODbin > 0] <- 1

deltaODbin <- deltaODbin %>%
  pivot_longer(everything(), names_to = "medium", values_to = "yngrow") 
deltaODbin$medium <- factor(deltaODbin$medium, 
                       levels = c('difcoMB', 'HMBcmpt', 'HMBpep', 'HMBaa', 'HMBlips', 'HMBoligo', 
                                  'HMBorg', 'HMBntrl', 'HMBamisug', 'HMBacdsug', 'HMB--'))

deltaODbin %>%
  ggplot(aes(x = medium, y = yngrow)) +
  geom_col() +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5),
        text = element_text(size = 14)) +
  labs(y = "Number of growing strains",
       x = "Medium")
```
